---
title: "Detroit Blight Violations Analysis (2022-2024)"
author: "Hamas Khanzada"
execute:
  echo: false
format: html
---

```{css}
/* Custom styling for tables and maps */
.hidden-content { display: none; }
.visible-content { display: block; }

table {
  width: 100%;
  border-collapse: collapse;
  margin: 1em 0;
}

.map-container {
  height: 600px;
  margin-bottom: 2em;
}

h2 {
  border-bottom: 1px solid #ddd;
  padding-bottom: 0.3em;
}
```

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(mapview)
library(lubridate)
library(viridisLite)
library(kableExtra)

# Set mapview options
mapviewOptions(fgb = FALSE, viewer.suppress = FALSE)
```

```{r, include=FALSE}
## Data Prepration
## Load Spatial Boundaries
#| message: false
#| warning: false

# Detroit city boundary
detroit_boundary <- st_read("data-raw/City_of_Detroit_Boundary.shp", quiet = TRUE) |>
  st_transform(4326)

# 2020 Census tracts
tracts_2020 <- st_read("data-raw/CensusTracts2020.shp", quiet = TRUE) |>
  st_transform(4326)  # Convert to WGS84 (lat/lon)

## Process Blight Violation Data
## Load and filter blight violations
blight_points <- read_csv("data-raw/Blight_Violations_20250228.zip",
                         col_types = cols(
                           ticket_id = col_character(),
                           violation_street_number = col_character(),
                           violation_zip_code = col_character(),
                           longitude = col_double(),
                           latitude = col_double(),
                           violation_date = col_character()
                         )) |>
  filter(!is.na(latitude) & !is.na(longitude)) |>
  mutate(violation_date = ymd_hms(violation_date)) |>
  filter(between(violation_date, as.Date('2022-01-01'), as.Date('2024-12-31')))

# Convert to spatial object
blight_points_sf <- st_as_sf(blight_points, 
                            coords = c("longitude", "latitude"),
                            crs = 4326)

## Spatial Analysis
## Count Violations Per Census Tract
# Spatial join and count
tract_counts <- tracts_2020 |>
  st_join(blight_points_sf) |>
  as_tibble() |>
  count(GEOID, name = "violation_count")

# Calculate area metrics
tract_blight <- tracts_2020 |>
  mutate(area_sq_mi = as.numeric(st_area(geometry)) / 27878400) |>
  left_join(tract_counts, by = "GEOID") |>
  mutate(
    violation_count = replace_na(violation_count, 0),
    violations_per_sq_mi = violation_count / area_sq_mi
  )
```
## Interactive Map
```{r}
#| warning: false
#| fig-height: 8

mapview(tract_blight, 
        zcol = "violation_count",
        layer.name = "Blight Violations (2022-2024)",
        col.regions = viridis(100),
        alpha.regions = 0.7,
        label = paste(
          "Tract:", tract_blight$GEOID, "<br>",
          "Total violations:", tract_blight$violation_count, "<br>",
          "Per sq mi:", round(tract_blight$violations_per_sq_mi, 1)
        )) +
  mapview(detroit_boundary, color = "red", lwd = 2)
```

## Background

What Are Blight Violations?

Blight violations are municipal citations for property conditions that violate local maintenance codes, including:

- Structural deterioration
- Overgrown vegetation
- Accumulated trash/debris
- Broken windows or graffiti

In Detroit, these violations reflect both property neglect and systemic challenges in neighborhood maintenance.

## Key Impacts

High blight violation density correlates with:

- ↓ Property values (15-30% reduction in affected areas)
- ↑ Crime rates (Broken Windows Theory effect)
- Public health risks (pest infestations, mold exposure)
- Municipal costs ($50M+ annually for enforcement/cleanup)

## About This Data:

```{r blight-data-sources, echo=FALSE}
library(kableExtra)  # Ensure this is loaded

data_sources <- tibble(
  `Data Element` = c("Blight Violations", "Census Tract Boundaries", "Detroit Boundary"),
  Description = c("Dataset containing measures for blight violations",
                 "2020 Census geometries for spatial analysis",
                 "Official city limits for mapping context"),
    Source = c(
    '<a href="https://data.detroitmi.gov/datasets/detroitmi::blight-violations/about">Blight Violations Dataset</a>',
    '<a href="">US Census Bureau</a>', 
    '<a href="https://data.detroitmi.gov/datasets/detroitmi::city-of-detroit-boundary/about">Detroit Boundary Dataset</a>'
  )
)

data_sources %>%
  kable(format = "html", escape = FALSE) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(2, width = "25em") %>%
  footnote(
    general = "Violation data filtered to 2022-2024 period with valid coordinates",
    general_title = "Note:"
  )
```

```{r}
## Summary Statistics
#| echo: false

tibble(
  Metric = c("Total Violations", "Average per Tract", "Max in Single Tract"),
  Value = c(
    sum(tract_blight$violation_count),
    mean(tract_blight$violation_count) |> round(1),
    max(tract_blight$violation_count)
  )
) |>
  kable(format = "html") |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```
## Methodology

1. Data Processing:
- Filtered violations to last three calendar years (2022-2024)
- Removed records missing location data
- Standardized all spatial data to WGS84 (EPSG:4326)

2. Spatial Analysis:
- Counted violations within each census tract using spatial join
- Calculated violation density per square mile

3. Visualization:
- Choropleth map shows raw counts
- Hover labels display density metrics
- Detroit boundary provides spatial context