---
title: "HRI-Mapping-Detroit_2020-Census_Tracts"
author: "Hamas Khanzada"
execute:
  echo: false
format: html
---

```{css}
/* Custom styles */
.hidden-content { display: none; }
.visible-content { display: block; }

table {
  width: 100%;
  border-collapse: collapse;
  margin: 1em 0;
}

h2 {
  border-bottom: 1px solid #ddd;
  padding-bottom: 0.3em;
}
```

```{r setup, include=FALSE}
#| message: false
library(tidyverse)
library(sf)
library(mapview)
library(leafem)
library(knitr)
library(kableExtra)
mapviewOptions(fgb = FALSE, viewer.suppress = FALSE)
```

```{r}
# Load and process data
HRI_data <- read.csv("data-raw/Historic_Redlining_Indicator_Detroit_2020.csv") %>%
  mutate(GEOID = as.character(GEOID))

detroit_tracts <- st_read("data-raw/CensusTracts2020.shp", quiet = TRUE)
detroit_boundary <- st_read("data-raw/City_of_Detroit_Boundary.shp", quiet = TRUE)

# Create joined dataset
detroit_joined <- detroit_tracts %>%
  left_join(HRI_data, by = "GEOID")
```
## Interactive Map
```{r}
#| fig-height: 8
red_palette <- colorRampPalette(c("#2166ac", "#b2182b"))

mapview(detroit_joined,
        zcol = "HRI2020",
        layer.name = "Redlining Impact<br>(Higher = Worse)",
        col.regions = red_palette,
        at = seq(1, 4, by = 0.5),
        alpha.regions = 0.7,
        lwd = 0.5,
        color = "white",
        label = paste("Tract:", detroit_joined$GEOID, "<br>Score:", detroit_joined$HRI2020)) + 
  mapview(detroit_boundary,
          color = "black",
          lwd = 2,
          alpha.regions = 0)
```
## Background

What is Redlining?

Redlining was a discriminatory practice begun in the 1930s where banks and federal agencies denied mortgages to residents in predominantly Black neighborhoods, marking them as "hazardous" on maps with red lines.

## Key Impacts

- Created persistent racial wealth gaps  
- Limited housing in minority communities  
- Shaped neighborhood disparities

## About This Data

The Historic Redlining Indicator (HRI) combines historical redlining patterns with contemporary socioeconomic data:    
```{r data-sources, echo=FALSE}
data_sources <- tibble(
  `Data Element` = c("HRI 2020 Scores", "Census Tract Boundaries", "City Boundary"),
  Description = c("Composite measure of historic redlining impact (1-4 scale)", 
                 "2020 Census tract geometries with GEOID matching",
                 "Official Detroit city limits"),
  Source = c(
    '<a href="https://www.openicpsr.org/openicpsr/project/141121/version/V3/view">HRI Dataset</a>',
    '<a href="https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html">US Census Bureau</a>',
    '<a href="https://data.detroitmi.gov/datasets/detroitmi::city-of-detroit-boundary/about">Detroit Boundary Dataset</a>'
  )
)

data_sources %>%
  kable(format = "html", escape = FALSE) %>%  # Ensure escape = FALSE
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE,
                font_size = 14) %>%
  column_spec(2, width = "25em") %>%
  footnote(general = "HRI scores combine 1930s HOLC grades with contemporary socioeconomic data",
           general_title = "Note:")
```
## Interpretation Guide
Understanding the Scores in Dataset:

A numerical value was assigned to each HOLC risk category as follows: 1 for “Least Impact”, 2 for “Moderate Impact”, 3 for “High Impact”, and 4 for “Severe Impact”. A historic redlining score was calculated from the summed proportion of HOLC residential security grades multiplied by a weighting factor based on area within each census tract. A higher score means greater redlining of the census tract.

```{r echo=FALSE}
tibble(Score = 1:4,
       Meaning = c("Least impact", "Moderate impact", "High impact", "Severe impact"),
       Color = c("#2166ac", "#67a9cf", "#ef8a62", "#b2182b")) %>%
  mutate(Color = cell_spec("", background = Color)) %>%  # Convert hex codes to color blocks
  kable(format = "html", escape = FALSE, align = "c") %>%
  kable_styling(full_width = FALSE, font_size = 14) %>%
  column_spec(3, width = "3em")  # Adjust width for color blocks
```
## Methodology
Data Processing Steps

1. Cleaned original HRI 2020 Census Tracts file by extracting census tracts only for Detroit by matching GeoIDs

2. Joined 2020 census tracts with HRI scores

2. Normalized all geographic data to NAD83 (EPSG:4269)

3. Visualized using Mapview
```{r}
#| classes: .small-code
glimpse(detroit_joined)
```